---
title: "visualization 1"
author: "Nhu Nguyen"
date: "2023-10-05"
output: github_document
---
loading necessary libraries 
```{r}
library(tidyverse)
library(ggridges)
```

## loading the weather data

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())

weather_df
```

## scatterplots!! 

create my first scatterplot ever 

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()
```

new approach, same plot 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()
```

save and edit a plot object

```{r}
weather_plot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax))

weather_plot + geom_point()
```

## advanced scatterplot

start with the same one and make it fancy! 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point() + 
  geom_smooth (se = FALSE)
```

what about the `aes` placement? 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name)) + 
    geom_smooth()
```

let's facet some things!! 

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```
facet function - '.' means don't create rows, but create columns that describe name variable 
alhpa refers to the transparency of the points, can also change size 

let's combine some elements and try a new plot 

```{r}
weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp), alpha = 0.5) + 
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

##some small notes 

how many geoms have to exist? 

you can have whatever geoms you want - may not be the most useful though if you only have the geom_smooth function

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_smooth(se = FALSE)
```

you can use  neat geom! 

geom_hex() gives me a 2D histogram, where brighter colors indicate more data points -- overkill for a data set with 1,000+ observations
  - helpful for understanding what the data distribution might look like 
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()
```

geom_bin2d makes the histogram points into squares 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_bin2d()
```
contour plot -- similar to an elevation map 
  - helpful when you have a large data set 
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_density2d() + 
  geom_point(alpha = .3)
```

## univarate plots 

histograms are really great 

only looking at one variable 
```{r}
weather_df %>% 
  ggplot(aes(x = tmin)) +
  geom_histogram()
```

can we add color? 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, fill = name)) +
  geom_histogram() + 
  facet_grid(. ~ name)
```

histograms might not be the best graphs to show distribution of one variable 

let's try a new geometry 
  - density = histogram that's been smooth out around the edges 
  - loose a little bit of information though but makes it easier to compare shapes of distribution 
    - good for exploration of data set 
```{r}
weather_df %>% 
  ggplot(aes( x = tmin, fill = name)) + 
  geom_density(alpha = .4)
```

what about box plots? 

```{r}
weather_df %>% 
  ggplot(aes(y = tmin)) + 
  geom_boxplot()
```
  - gives one box plot of the overall distribution of tmin 
    - the median, min, max of the temperature mins
    
if we want to seperate by `name`: 
  - strange because we are putting a variable that is not a number on the x-axis 
  - however, since it is a box plot, it works out ok 

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmin)) + 
  geom_boxplot()
```

trendy plots :-) 

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmin, fill = name)) +
  geom_violin(alpha = .5) +
  stat_summary(fun = "median", color = "blue")
```
  - violin plots: a cross mix of box plot and density plot 
  - this is more informative than box-plot about distribution shape 

ridge plots -- most popular plot of 2017
  - when you want the density/distribution on the x-axis and y is a predictor 
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = name)) +
  geom_density_ridges()
```
  - ridge plots are easier when you have a categorical predictor with 40+ variables 
  - ex: if you wanted to know what the age distribution looked like according to states, using box-plots or violin plots would be difficult to graph -- ridge plots end up being a useful option 
  
## save and embed

let's save a scatterplot 

```{r}
weather_plot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5)

ggsave("./results/weather_plot.pdf", weather_plot, width = 8, height = 5)
```

what about embedding?
  - are there ways we can control what specific figures look like? 
  
```{r, fig.width= 6, fig.height= 3}
weather_plot
```

embed at different size 
```{r, fig.width= 12, fig.height= 6}
weather_plot
```

  - made same plot, but it looks different when we knit -- WHY? 
    - first plot: we made the figure smaller but the points are the still same
    
we can also just set it so that every graph will have the same dimensions

```{r}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name))
```


