---
title: "visualization 2"
author: "Nhu Nguyen"
date: "2023-10-05"
output: github_document
---
loading necessary libraries 
```{r}
library(tidyverse)
library(patchwork)
```

## loading the weather data

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())

weather_df
```

## remember this plot? 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5)
```
this plot is good for me to know what is going on in the graph, but if someone else were to look at this graph, they might not know what tmin or tmax is -- might be helpful to have a caption 

## labels

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) + 
  labs(
    title = "temperature plot", 
    x = "minimum daily temperature (x)",
    y = "maximum daily temperature (y)", 
    color = "location", 
    caption = "data from the rnoaa package: temperatures in 2017"
  )
```

## scales 

start with the same plot, x and y scale

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) + 
  labs(
    title = "temperature plot", 
    x = "minimum daily temperature (x)",
    y = "maximum daily temperature (y)", 
    color = "location", 
    caption = "data from the rnoaa package: temperatures in 2017"
  ) + 
scale_x_continuous(
  breaks = c(-15, 0, 15),
  labels = c("-15ยบ C", "0", "15")
) + 
scale_y_continuous(
  trans = "sqrt"
)
```

look at color scales

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) + 
  labs(
    title = "temperature plot", 
    x = "minimum daily temperature (x)",
    y = "maximum daily temperature (y)", 
    color = "location", 
    caption = "data from the rnoaa package: temperatures in 2017"
  ) + 
  scale_color_hue(h = c(100, 300))
```


```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) + 
  labs(
    title = "temperature plot", 
    x = "minimum daily temperature (x)",
    y = "maximum daily temperature (y)", 
    color = "location", 
    caption = "data from the rnoaa package: temperatures in 2017"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  )
```

## themes 

shift the legend 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) + 
  labs(
    title = "temperature plot", 
    x = "minimum daily temperature (x)",
    y = "maximum daily temperature (y)", 
    color = "location", 
    caption = "data from the rnoaa package: temperatures in 2017"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE) + 
  theme(legend.position = "bottom")
```

change the overall theme

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) + 
  labs(
    title = "temperature plot", 
    x = "minimum daily temperature (x)",
    y = "maximum daily temperature (y)", 
    color = "location", 
    caption = "data from the rnoaa package: temperatures in 2017"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE) + 
  theme_minimal() + 
  theme(legend.position = "bottom")
```

## setting options

  - put at the very beginning of rmarkdown file -- makes sure that every plot i make has the viridis color palate, the theme will be minimal, and the legend will be at the bottom 
  
```{r, eval = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6, 
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis", 
  ggplot2.continuous.fill = "viridis"
)

scale_color_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## data args in `geom`

```{r}
central_park_df = 
  weather_df |> 
  filter(name == "CentralPark_NY")

molokai_df = 
  weather_df |> 
  filter(name == "Molokai_HI")

ggplot(data = molokai_df, aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_line(data = central_park_df)
```

##  `patchwork` 

remember faceting? 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, fill = name)) + 
  geom_density(alpha=.5) + 
  facet_grid(. ~ name)
```

what happens when you want multipanel plots but can't facet?     - where `patchwork` comes in

```{r}
tmax_tmin_p =
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  theme(legend.position = "none")

prcp_dens_p = 
  weather_df %>% 
  filter(prcp > 0) %>% 
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .5) +
  theme(legend.position = "none")

tmax_date_p = 
  weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(alpha = .5) + 
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom")

(tmax_tmin_p + prcp_dens_p) / tmax_date_p
```

## data manipulation 

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) + 
  geom_violin(alpha = .5)
```
  - for character variables, R turns the names into factors (i.e. centralpark = 1) and then puts it in alphabetical order     - in order to put the labels in a different order, need to manipulate my data by using factors -- not a ggplot problem 
  - mutate has turned the variable from character to factor 
  
control your factors
  
```{r}
weather_df %>% 
  mutate(
    name = factor(name), 
    name = forcats::fct_relevel(name, c("Molokai_HI"))
  ) %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) + 
  geom_violin(alpha = .5) 
```

what if i wanted densities for tmin and tmax simultaneously? 

```{r}
weather_df |>
  pivot_longer(
    tmax:tmin,
    names_to = "observation", 
    values_to = "temp"
  ) %>% 
  ggplot(aes(x = temp, fill = observation)) +
  geom_density(alpha = .5) +
  facet_grid(. ~ name)
```




